A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
rework on variable transformation and scaling

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|--------------|----------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

@@

`Output`: $y^l(s; t)$

`Computation`: The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

```{=tex}
\begin{equation}
h_{\tau}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$. Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.

<!-- A common assumption is normality in drought index.  -->

@@

`Output`: $x^l(s; t)$

`Computation`: While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline. The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of

```{=tex}
\begin{equation}
[x(s_i;t_j) - \alpha]/\gamma.
\end{equation}
```
For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$. @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.

<!-- Take one or multiple variables to create a new variable,  -->

<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->

<!-- change the shape of the variable -->

<!-- <!-- A conversion of non-normal variables into (approximately) normal. wrong! -->

<!-- GAM, can you do additive model pairwise/ three-way -->

<!-- While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect. -->

```{r fig-scale-var-trans-compare}
#| fig-cap: "Comparison of operations in scaling (green) and variable transformation (orange) steps in free scale. Variables after the scaling operations have the same distribution as the origin, while the distribution changes after variable transformation. "
#| fig-height: 8
#| fig-width: 12
set.seed(123)
x <- rgamma(n = 1000, shape = 1, scale = 2)

@@

latex_df <- tibble(
  var = c("origin", "z-score", "qudratic", "log", "square root", "cubic root",
          "minmax", "boxcox", "centering"),
  latex = c(
           r"($x$)",
           r"($\frac{x - \bar{x}}{sd(x)}$)",
           r"($x^2$)",
           r"($log(x)$)",
           r"($\sqrt{x}$)",
           r"($x^{1/3}$)",
           r"($\frac{x - min(x)}{max(x) - min(x)}$)",
           r"($\frac{x^{0.5} - 1}{0.5}$)",
           r"($x - \bar{x}$)"
         )
) %>% 
  left_join(trans_df %>% group_by(var) %>% summarise(max = max(value)))

trans_df %>% 
  ggplot() +
  geom_density(aes(x = value, color = category, y = after_stat(scaled))) + 
  geom_label(data = latex_df, 
             aes(x = 0.8 * max, y = 0.8, label = TeX(latex, output = "character")), 
             parse = TRUE) + 
  facet_wrap(vars(var), scales = "free") + 
  scale_color_brewer(palette = "Dark2") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

>>>>

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|--------------|----------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x_t(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(s;t) = g[x(s;t)]$. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

@@

Variable transformation takes input of a single variable $x^i(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^i(s;t)$. When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality. For example, in Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional shape of the variable. It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x^i(s;t) - \alpha]/\gamma$. In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.

```{r fig-scale-var-trans-compare}
#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange). While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation. "
#| fig-height: 8
#| fig-width: 12
set.seed(123)
x <- rgamma(n = 1000, shape = 1, scale = 2)

@@

latex_df <- tibble(
  var = c("origin", "z-score", "qudratic", "log", "square root", "cubic root",
          "minmax", "boxcox", "centering"),
  latex = c(
           r"($x$)",
           r"($\frac{x - \bar{x}}{sd(x)}$)",
           r"($x^2$)",
           r"($log(x)$)",
           r"($\sqrt{x}$)",
           r"($x^{1/3}$)",
           r"($\frac{x - x_{min}}{x_{max} - x_{min}}$)",
           r"($\frac{x^{0.5} - 1}{0.5}$)",
           r"($x - \bar{x}$)"
         )
) %>% 
  left_join(trans_df %>% group_by(var) %>% summarise(max = max(value)))

trans_df %>% 
  ggplot() +
  geom_density(aes(x = value, color = category, y = after_stat(scaled))) + 
  geom_label(data = latex_df, 
             aes(x = 0.8 * max, y = 0.8, label = TeX(latex, output = "character")), 
             parse = TRUE) + 
  facet_wrap(vars(var), scales = "free") + 
  scale_color_brewer(palette = "Dark2", name = "module") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

### 2023-08-15 tidyindex.tex
rework on variable transformation and scaling

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
\(Tx^i(s;t)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where
\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p^i(s;t)\), where
\(p(.) \in [0, 1]\) is a probability value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
\(z^i(s; t)\), where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
hyper-parameters associated with the transformation. The change of
spatial dimension allows for the alignment of variables collected from
different measurements, such as in-situ stations and satellite imagery,
or originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
countries scales.

@@

\texttt{Output}: \(y^l(s; t)\)

\texttt{Computation}: The purpose of variable transformation is to
create variables that fits assumptions for further computing. These
assumptions include a stable variance, normal distribution, or a certain
scale required by some algorithms down the pipeline. Variable
transformation is a general notion of a functional transformation on the
variable:

\begin{equation}
h_{\tau}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in
the transformation if any, and \(d_{\tau}\) is the number of parameter
of \(\tau\). Transformation is needed for data that are highly skewed
and some common transformations include log, quadratic, and square root
transformation.

@@

\texttt{Output}: \(x^l(s; t)\)

\texttt{Computation}: While scaling can be seen as a specific type of
variable transformation, it is separated into its own step to make the
step explicit in the pipeline. The key difference between the two steps
is that variable transformation typically changes the shape of the data
while scaling only changes the data scale and can usually be written in
the form of

\begin{equation}
[x(s_i;t_j) - \alpha]/\gamma.
\end{equation}

For example, a z-score standardisation can be written in the above form
with \(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\), a min-max
standardisation uses \(\alpha = \min[x(s_i, t_j)]\) and
\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
pre-processing operations and uses color to differentiate whether the
operation is a variable transformation or a scaling step. While both
variable transformation and scaling are pre-processing steps, the
scaling operations in green show the same distribution as the original
data.

@@

\caption{\label{fig-scale-var-trans-compare}Comparison of operations in
scaling (green) and variable transformation (orange) steps in free
scale. Variables after the scaling operations have the same distribution
as the origin, while the distribution changes after variable
transformation.}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
\(Tx^i(s;t)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where \(F\) is
the CDF function & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a
probability value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
\(z^i(s; t)\), where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x_t(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(s;t) = g[x(s;t)]\). The change of spatial dimension allows for
the alignment of variables collected from different measurements, such
as in-situ stations and satellite imagery, or originating from different
resolutions. This also includes the aggregation of variables into
different levels, such as city, state, and countries scales.

@@

Variable transformation takes input of a single variable \(x^i(s;t)\)
and reshapes its distribution using the function \(T[.]\) to produce
\(x^i(s;t)\). When a variable has a skewed distribution, transformations
such as log, square root, or cubic root can adjust the distribution
towards normality. For example, in Human Development Index (HDI), a
logarithmic transformation is applied to the variable Gross National
Income per capita (GNI), to reduce its impact on HDI, particularly for
countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardisation, and min-max standardisation and can be expressed as
\([x^i(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
three dimensions (health, education, and economy) are converted into the
same scale (0-1) using min-max standardisation.

@@

\caption{\label{fig-scale-var-trans-compare}Comparison of the module
scaling (green) and variable transformation (orange). While both modules
change the variable range, scaling maintains the same distributional
shape, which is not the case with variable transformation.}
